---
title: "Skeleton Clustering : A Dimension free Density-Aided Clustering"
author: "| Kaustav Paul\n| Sourav Biswas \n"
institute: "Indian Statistical Institute, Kolkata"
date: "`r Sys.Date()`"
output: 
  beamer_presentation:
    theme: "AnnArbor"
    colortheme: "beaver"
---

```{r setup, include=FALSE, comment=NA}
knitr::opts_chunk$set(echo = FALSE)
```
## Traditional Clustering Methods



## Skeleton Clustering Framework.

**Input :** Observations $X_1,X_2,\ldots,X_N$, final number of clusters $S$.

1.  **Knot construction** : Perform $k-$means clustering with a large number $k$; the centers are the knots.
2.  **Edge construction** : Apply approximate Delaunay triangulation to the knots. Generally we choose $k=\left\lfloor\sqrt{n}\right\rfloor$
3.  **Edge weights construction** : Add weights to each edge using either Voronoi density, Face density or Tube density similarity measure.
4.  **Knots segmentation** : Use linkage criterion to segment knots into $S$ groups based on the edge weights.
5.  **Assignment of labels** : Assign a cluster label to each observation based on which knot group the nearest knot belongs to.

## Knot construction

-   Some knots are constructed to give a concise representation of the data structure.
-   In practice we use $k$-Means to choose $k=\left\lfloor\sqrt{n}\right\rfloor$ knots, where $n$ is the number of samples.
-   Empirically robustness performance with sufficient number of knots.

```{r}

```











